Refinement of Thermostated Molecular Dynamics Using Backward Error Analysis

### 2018-11-10 rigid_bodies.tex
Further changes in the discretization error correction section

\newenvironment{smallarray}[1]                          % small arrays
{
	\null\,\vcenter\bgroup\scriptsize
	\renewcommand{\arraystretch}{1.5}
	\arraycolsep=.13885em
	\hbox\bgroup$\left[\array{@{}#1@{}}}
	{\endarray\right]$\egroup\egroup\,\null
}

@@

\title{Refinement of Thermostated Molecular Dynamics via Backward Error Analysis}

@@

Debenedetti....

@@

In these equations, $\vt r$ is the center-of-mass position of the body, $\vt p$ is its linear momentum, $\vt q$ is a unit quaternion that determines its orientation, and $\vt \pi$ is the quaternion-conjugate momentum.
Vectors $\vt F$ and $\vt \tau$ are, respectively, the resultant force and torque exerted on the body, both represented in the space-fixed frame of reference, and $\vt \omega = \frac{1}{2} {\mt I}^{-1} \tr{\mt B}(\vt q) {\vt \pi}$ is the three-dimensional angular velocity in the body-fixed frame.
Matrices $\mt M$ and $\mt I$ are diagonal ones.
The former contains the body mass in all three diagonal entries while the latter contains the three principal moments of inertia.
Finally, the matrix-valued functions $\mt B$ and $\vt C$, given by
\begin{equation*}
\label{eq:def_B_and_C}
\mt B(\vt q) = \begin{smallarray}{rrrr}
-q_2 & -q_3 & -q_4 \\
 q_1 & -q_4 &  q_3 \\
 q_4 &  q_1 & -q_2 \\
-q_3 &  q_2 &  q_1
\end{smallarray}
%
\; \text{and} \;
%
\mt C(\vt q) = \begin{smallarray}{rrrr}
-q_2 & -q_3 & -q_4 \\
 q_1 &  q_4 & -q_3 \\
-q_4 &  q_1 &  q_2 \\
 q_3 & -q_2 &  q_1
\end{smallarray},
\end{equation*}
are related to the rotation matrix ${\mt A}(\vt q)$ through the factorization ${\mt A}(\vt q) = \tr{\mt B}(\vt q) {\mt C}(\vt q)$ that was mentioned at the beginning of this section. Eq.~\eqref{eq:ODE system for NVE} preserves the constraints $\tr{\vt q}{\vt q} = 1$ and $\tr{\vt q}{\vt \pi} = 0$, as well as the value of a Hamiltonian
\begin{equation}
\label{eq:Hamiltonian}
\Ham{} = K(\vt p, \vt q, \vt \pi) + U(\vt r,\vt q),
\end{equation}
where $K$ and $U$ are the kinetic and potential energies of the body, respectively.
While $U(\vt r, \vt q)$ is an arbitrary function, the kinetic energy is given by
\begin{equation}
\label{eq:kinetic energy}
K = \frac{1}{2} \tr{\vt p} {\mt M}^{-1} {\vt p} + \frac{1}{8} \tr{\vt \pi} {\mt B}(\vt q) {\mt I}^{-1} \tr{\mt B}(\vt q) {\vt \pi}.
\end{equation}

@@

Regarding the refinement of stochastic methods, the results obtained from the ``Refined CSVR'' scheme are similar to those from other refined schemes.
This was expected because the unrefined versions of the CSVR \cite{Bussi_2007} (stochastic) and the NHC \cite{Martyna_1996} (deterministic) thermostats for rigid bodies produce virtually identical thermodynamic properties of water as functions of the time-step size (results not shown).
Similar behavior is expected for other global thermostats such as the Nos\'{e}-Hoover-Langevin method \cite{Samoletov_2007, Leimkuhler_2009}.
Extension of our approach to massive Langevin-type thermostats \cite{Brunger_1984, Davidchack_2009, Davidchack_2015}, in which every degree of freedom is acted upon independently, is beyond the scope of this work.
Nonetheless, we envision that such an extension would entail using the components of entrywise products between $\vt p$ and $\dot{\vt r}^{[n]}$, and also between ${\vt \pi}$ and $\dot{\vt q}^{[n]}$ in the case of rigid bodies, instead of the scalar products present in Eq.~\eqref{eq:modified kinetic energy estimator}.

Interestingly, all thermostatting schemes analyzed in Fig.~\ref{fig:properties}, including the KLN method, yield similar results for $c_\mini{V}$.
Even though the average $\langle \Ham{} \rangle$ gets affected by discretization errors, Fig.~\ref{fig:properties}(d) shows that the magnitude of fluctuations in the sampled $\Ham{}$ values does not, and this is successfully predicted by all schemes.
To put it differently, consider that discretization errors in a simulation at temperature $T$ cause the computed average $\langle \Ham{} \rangle$ to erroneously correspond to a nearby temperature, say, $T^\ast$.
If the specific heat of the system changes little with temperature in the proximities of $T$, then it is natural to expect little effect of the same discretization errors on the computed $c_\mini{V}$.
This is the case of liquid water in the vicinities of $298~\mathrm{K}$.

With the aim of highlighting the importance of reweighting, in Fig.~\ref{fig:free_energy} we show for the ``Refined NHC'' scheme the free energy difference per water molecule ($\Delta F/N$) between the sampled and target states as a function of the time-step size.
$\Delta F$ corresponds to the logarithm of the averaged denominator of Eq.~\eqref{eq:reweighting}.
It should be noted that the other refined schemes yield identical results and, as expected, the free energy is only negligible for $h = 1~\text{fs}$.

@@

A significant enhancement was achieved for a system of TIP3P rigid water molecules\cite{Jorgensen_1983} by using an analytically approximation to the Shadow Hamiltonian (here refered to as \textit{refined Hamiltonian}) with a minimal computatioal overhead, and whose derivation was greatly facilitated by the unsplit solution of rotation introduced in our previous paper\cite{Silveira_2017}.
Particularly, we have been able to remove the serious artifacts observed in the schemes of Martyna \textit{et al}. \cite{Martyna_1996} and Kamberaj \textit{et al}. \cite{Kamberaj_2005}, correcting the discretization effects in numerically computed averages for time-step sizes up to $5~\text{fs}$.
From our rigorous analysis of these schemes, we would not recommend a time-step size larger than $1~\text{fs}$.
The fact of having overcome that barrier is of paramount importance as allows us to actually exploit the high numerical stability provided by rigid-body MD.

>>>>

@@

\title{Refinement of Thermostated Molecular Dynamics Using Backward Error Analysis}

@@

Debenedetti.... \cite{Palmer_2018} \cite{Limmer_2011, Limmer_2013}

@@

In these equations, $\vt r$ is the center-of-mass position of the body, $\vt p$ is its linear momentum, $\vt q$ is a unit quaternion that determines its orientation, and $\vt \pi$ is the quaternion-conjugate momentum.
Vectors $\vt F$ and $\vt \tau$ are, respectively, the resultant force and torque exerted on the body, both represented in the space-fixed frame of reference, and $\vt \omega = \frac{1}{2} {\mt I}^{-1} \tr{\mt B}(\vt q) {\vt \pi}$ is the three-dimensional angular velocity in the body-fixed frame.
Matrices $\mt M$ and $\mt I$ are diagonal ones.
The former contains the body mass in all three diagonal entries while the latter contains the three principal moments of inertia.
Finally, the matrix-valued functions $\mt B$ and $\vt C$, given by
\begin{equation*}
\label{eq:def_B_and_C}
\mt B(\vt q) = \left[\begin{array}{rrrr}
-q_2 & -q_3 & -q_4 \\
 q_1 & -q_4 &  q_3 \\
 q_4 &  q_1 & -q_2 \\
-q_3 &  q_2 &  q_1
\end{array}\right]
\end{equation*}
and
\begin{equation*}
\mt C(\vt q) = \left[\begin{array}{rrrr}
-q_2 & -q_3 & -q_4 \\
 q_1 &  q_4 & -q_3 \\
-q_4 &  q_1 &  q_2 \\
 q_3 & -q_2 &  q_1
\end{array}\right]
\end{equation*}
are related to the rotation matrix ${\mt A}(\vt q)$ through the factorization ${\mt A}(\vt q) = \tr{\mt B}(\vt q) {\mt C}(\vt q)$ that was mentioned at the beginning of this section. Eq.~\eqref{eq:ODE system for NVE} preserves the constraints $\tr{\vt q}{\vt q} = 1$ and $\tr{\vt q}{\vt \pi} = 0$, as well as the value of a Hamiltonian
\begin{equation}
\label{eq:Hamiltonian}
\Ham{} = K(\vt p, \vt q, \vt \pi) + U(\vt r,\vt q),
\end{equation}
where $K$ and $U$ are the kinetic and potential energies of the body, respectively.
While $U(\vt r, \vt q)$ is an arbitrary function, the kinetic energy is given by
\begin{equation}
\label{eq:kinetic energy}
K = \frac{1}{2} \tr{\vt p} {\mt M}^{-1} {\vt p} + \frac{1}{8} \tr{\vt \pi} {\mt B}(\vt q) {\mt I}^{-1} \tr{\mt B}(\vt q) {\vt \pi}.
\end{equation}

@@

Regarding the refinement of stochastic methods, the results obtained from the ``Refined CSVR'' scheme are similar to those from other refined ones.
This was expected because the unrefined versions of the CSVR \cite{Bussi_2007} (stochastic) and the NHC \cite{Martyna_1996} (deterministic) thermostats for rigid bodies produce virtually identical properties as functions of the time-step size (results not shown).
Similar behavior is expected for other global thermostats such as the Nos\'{e}-Hoover-Langevin method \cite{Samoletov_2007, Leimkuhler_2009}.
Extension of our approach to massive Langevin-type thermostats \cite{Brunger_1984, Davidchack_2009, Davidchack_2015}, in which every degree of freedom is acted upon independently, is beyond the scope of this work.
Nonetheless, we envision that this would entail using the components of entrywise products between $\vt p$ and $\dot{\vt r}^{[n]}$, and also between ${\vt \pi}$ and $\dot{\vt q}^{[n]}$ in the case of rigid bodies, instead of the scalar products present in Eq.~\eqref{eq:modified kinetic energy estimator}.

Fig.~\ref{fig:properties}(d) contains results for the constant-volume specific heat capacity.
Interestingly, these results do not get affected by the discretization errors when $\timestep \leq 5~\mathrm{fs}$, not even if the KLN method is employed.
Given that the specific heat of water changes little with temperature in the vicinity of $298~\mathrm{K}$, it is not surprising to observe that $c_\mini{V}$ remains unaltered while other properties vary.

With the aim of highlighting the importance of the reweighting procedure, in Fig.~\ref{fig:free_energy} we show the free energy difference per water molecule ($\Delta F/N$) between the actually sampled and the target states, obtained with the ``Refined NHC'' scheme as a function of the time-step size.
Note that $\Delta F$ corresponds to the logarithm of the denominator of Eq.~\eqref{eq:reweighting}.
It should be mentioned that the other refined schemes yield similar results and, as expected, the free energy difference is non-negligible except for the smallest time steps.

@@

A significant enhancement was achieved for a system of TIP3P rigid water molecules\cite{Jorgensen_1983} by using an analytically approximation to the Shadow Hamiltonian (here referred to as \textit{refined Hamiltonian}) with a minimal computational overhead, and whose derivation was greatly facilitated by the unsplit solution of rotation introduced in our previous paper\cite{Silveira_2017}.
Particularly, we have been able to remove the serious artifacts observed in the schemes of Martyna \textit{et al}. \cite{Martyna_1996} and Kamberaj \textit{et al}. \cite{Kamberaj_2005}, correcting the discretization effects in numerically computed averages for time-step sizes up to $5~\text{fs}$.
From our rigorous analysis of these schemes, we would not recommend a time-step size larger than $1~\text{fs}$.
The fact of having overcome that barrier is of paramount importance as allows us to actually exploit the high numerical stability provided by rigid-body MD.
